Symbolic On-the-fly Algorithms for GKAT Equivalences \thanks{Identify applicable funding agency here. If none, delete this.}

### 2024-12-31 main.tex
pass on universal coalgebra

\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{6\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
}

\maketitle

@@

TODO: change all the \(e_b\) to \(b\).

@@

Our notion of normalized bisimulation is inherently coalgebraic, thus it is empirical for us to recall some notions and theorems in coalgebra.
Given a functor \(F\) on the category of set and functions, a \emph{coalgebra over \(F\)} or \emph{\(F\)-coalgebra} consists of a set \(S\) and a function \(δ_S: S → F(S)\).
We typically call \(F\) the \emph{signature} of the coalgebra, \(S\) the states of the coalgebra, and \(δ_S\) the transition function of the coalgebra.
We will sometimes use the states \(S\) to denote the coalgebra, when no ambiguity can arise.

A homomorphism between two \(F\)-coalgebra \(S\) and \(T\) is a map \(h: S → T\) that preserves the transition function; diagrammatical, the following diagram commutes:
\[
    \begin{tikzcd}
        S \ar{r}{h} \ar[swap]{d}{δ_S} & T \ar{d}{δ_T} \\  
        F(S) \ar{r}{F(h)} & F(T)
    \end{tikzcd}    
\]

When we can restrict the homomorphism map into a inclusion map \(i: S' → S\) for \(S' ⊆ S\), then we say that \(S'\) is a \emph{sub-coalgebra} of \(S\), denoted as \(S' ⊑ S\). Specifically, the following diagram commutes when \(S' ⊑ S\):
\[
    \begin{tikzcd}
        S' \ar[hook]{r}{i} \ar[swap]{d}{δ_{S'}} & S \ar{d}{δ_S} \\  
        F(S') \ar[hook]{r}{F(i)} & F(S)
    \end{tikzcd}    
\]
In fact, the transition function \(δ_{S'}\) is uniquely determined by the states \(S'\)~\cite[Proposition 6.1]{rutten_UniversalCoalgebraTheory_2000}.

The sub-coalgebras are preserved under homomorphic images and pre-images: 
\begin{lemma}[Theorem 6.3~\cite{rutten_UniversalCoalgebraTheory_2000}]\label{thm:hom-(pre)img-preserve-sub-coalg}
    given a homomorphism \(h: S → T\), and sub-coalgebras \(S' ⊑ S\) and \(T' ⊑ T\), then 
    \[h(S') ⊑ T \text{ and } h^{-1}(T') ⊑ S.\]
\end{lemma}

One particularly important sub-coalgebra of and coalgebra \(S\) is the least coalgebra generated by a single element \(s\), we will typically denote it as \(⟨s⟩_{S}\), and call it \emph{principle sub-coalgebra} generated by \(s\). We will omit the subscript \(S\) when it can be inferred from context or irrelevant.
As we will see later, principle sub-coalgebra \(⟨s⟩_S\) is important because it models all the reachable state of \(s\) in \(S\).

Principle sub-coalgebra always exists, because sub-coalgebra of any coalgebra forms a complete lattice~\cite[theorem 6.4]{rutten_UniversalCoalgebraTheory_2000}.
And similar to sub-coalgebra, principle sub-coalgebra is also preserved under homomorphic image:
\begin{theorem}\label{thm:homo-img-preserve-principle-sub-coalg}
    Homomorphic image preserves principle sub-GKAT coalgebra. Specifically, given a homomorphism \(h: S → T\):
    \[h(⟨s⟩_S) = ⟨h(s)⟩_T\]
\end{theorem}

\begin{proof}
    We will need to show that \(h(⟨s⟩_{S})\) is the smallest sub-GKAT coalgebra of \(S\) that contain \(h(s)\). First by definition of image, \(h(s) ∈ h(⟨s⟩_{S})\); second by \cref{thm:hom-(pre)img-preserve-sub-coalg}, \(h(⟨S⟩_S) ⊑ T\).

    Finally, take any \(T' ⊑ T\) and \(h(s) ∈ T'\), recall that by \cref{thm:hom-(pre)img-preserve-sub-coalg}, \(h^{-1}(T') ⊑ S\). We can then derive that \(h(⟨s⟩_S) ⊑ T'\): 
    \begin{align*}
        h(s) ∈ T' 
        & ⟹ s ∈ h^{-1}(T') \\  
        & ⟹ ⟨s⟩_S ⊑ h^{-1}(T') & \text{definition of \(⟨s⟩_S\)}\\  
        & ⟹ h(⟨s⟩_S) ⊑ T' & \text{take the image of \(h\) and \cref{thm:hom-(pre)img-preserve-sub-coalg}}
    \end{align*}
    Hence \(h(⟨s⟩_S)\) is the smallest sub-GKAT coalgebra of \(T\) that contains \(h(s)\).
\end{proof}

@@

Given two \(F\)-coalgebra \(S\) and \(T\), the \emph{behavioral equivalence} between states in \(S\) and \(T\) can be computed by a notion called \emph{bisimulation}.
A relation \({∼} ⊆ S × T\) is called a \emph{bisimulation relation} if it forms a \(F\)-coalgebra: \[δ_{∼}: {∼} → F(∼),\] 
And its projection functions \(π₁: S × T → S\) and \(π₂: S × T → T\) are both homomorphisms:
\[
    \begin{tikzcd}
        S \ar[swap]{d}{δ_S}
            & ∼ \ar[swap]{l}{π₁} \ar{r}{π₂} \ar{d}{δ_{∼}} 
            & T \ar{d}{δ_T}\\  
        F(S) & F(∼) \ar{l}{F(π₁)} \ar[swap]{r}{F(π₂)} & F(T)
    \end{tikzcd}
\]
As we will see later, bisimulation or variation of bisimulation indeed correspond to important semantical equivalences.

\begin{theorem}
    Given a functionr \(F\) on set, two \(F\)-colagebra \(S\) and \(T\), if there exists a surjective homomorphism \(h: S → T\), then \(S\) and \(T\) are bisimular.
\end{theorem}

@@

    The condition \labelcref{itm:acc-condition} holds:
    \begin{align*}
        δ_S(s, α) = \accept 
        & ⟺ \norm(δ_S)(s, α) = \accept 
        ⟺ \norm(δ_∼)((s,t), α) = \accept \\ 
        & ⟺ \norm(δ_T)(t, α) = \accept 
        ⟺ δ_T(t, α) = \accept
    \end{align*}

    The condition~\labelcref{itm:transition-bisim} holds, by case analysis on the liveness of \(s'\) and \(t'\).
    First note that \(s'\) and \(t'\) has to be both live or both dead: because \(δ_S(s, α) = (s', p)\), then \(\norm(δ_S)(s', α)\) can either be rejection or \((s',p)\), and so is \(\norm(δ_T)(t', α)\). Finally because \(s ∼ t\), then
    \[s' \text{ is live} ⟺ \norm(δ_S)(s, α) = (s', p) ⟺ \norm(δ_T)(t, α) = (t', p) ⟺ t' \text{ is live}.\]
    \begin{itemize}
        \item If both \(s'\) and \(t'\) are live, then \(s' ∼ t'\). By~\cref{thm:sub-coalg-preserve-bisim}, the bisimulation \(∼_{s', t'}\) is just \(∼\) restricted to \(⟨s'⟩\) and \(⟨t'⟩\).
        \item If both \(s'\) and \(t'\) are dead, then \(∼_{s', t'}\) can just be the singleton relation, according to~\cref{thm:bisim-between-dead}.
    \end{itemize}

@@

    The condition~\labelcref{itm:rej-or-dead} holds: 
    \begin{align*}
        δ_S(s, α) \text{ rejects or transition to dead states} 
    & ⟺ \norm(δ_S)(s, α) = \reject \\
    & ⟺ \norm(δ_T)(t, α) = \reject \\
    & ⟺ δ_T(t, α) \text{ rejects or transition to dead states}.
    \end{align*}

>>>>

\author{\IEEEauthorblockN{Cheng Zhang\textsuperscript{\textsection}}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University College London}\\
London, United Kingdom \\
0000-0002-8197-6181}
\and
\IEEEauthorblockN{Qiancheng Fu}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
email address or ORCID}
\and
\IEEEauthorblockN{Hang Ji}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
email address or ORCID}
\and
\IEEEauthorblockN{Ines Santacruz}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
email address or ORCID}
\and
\IEEEauthorblockN{Marco Gaboardi}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
email address or ORCID}
}

\maketitle
\begingroup\renewcommand\thefootnote{\textsection}
\footnotetext{Work largely performed at Boston University}
\endgroup

@@

@@

In this paper, we will make heavy use of coalgebraic theory, thus it is empirical for us to recall some notions and useful theorems in universal coalgebra.
Given a functor \(F\) on the category of set and functions, a \emph{coalgebra over \(F\)} or \emph{\(F\)-coalgebra} consists of a set \(S\) and a function \(σ_S: S → F(S)\).
We typically call elements in \(S\) the \emph{states} of the coalgebra, and \(σ_S(s)\) the \emph{dynamic} of state \(s\).
We will sometimes use the states \(S\) to denote the coalgebra, when no ambiguity can arise.

A homomorphism between two \(F\)-coalgebra \(S\) and \(U\) is a map \(h: S → U\) that preserves the function \(σ\); diagrammatically, the following diagram commutes:
\[
    \begin{tikzcd}
        S \ar{r}{h} \ar[swap]{d}{σ_S} & U \ar{d}{σ_U} \\  
        F(S) \ar{r}{F(h)} & F(U)
    \end{tikzcd}    
\]

When we can restrict the homomorphism map into a inclusion map \(i: S' → S\) for \(S' ⊆ S\), then we say that \(S'\) is a \emph{sub-coalgebra} of \(S\), denoted as \(S' ⊑ S\). Specifically, the following diagram commutes when \(S' ⊑ S\):
\[
    \begin{tikzcd}
        S' \ar[hook]{r}{i} \ar[swap]{d}{σ_{S'}} & S \ar{d}{σ_S} \\  
        F(S') \ar[hook]{r}{F(i)} & F(S)
    \end{tikzcd}    
\]
In fact, the function \(σ_{S'}\) is uniquely determined by the states \(S'\)~\cite[Proposition 6.1]{rutten_UniversalCoalgebraTheory_2000}.

The sub-coalgebras are preserved under homomorphic images and pre-images: 
\begin{lemma}[Theorem 6.3~\cite{rutten_UniversalCoalgebraTheory_2000}]\label{thm:hom-(pre)img-preserve-sub-coalg}
    given a homomorphism \(h: S → U\), and sub-coalgebras \(S' ⊑ S\) and \(U' ⊑ U\), then 
    \[h(S') ⊑ U \text{ and } h^{-1}(U') ⊑ S.\]
\end{lemma}

One particularly important sub-coalgebra of and coalgebra \(S\) is the least coalgebra generated by a single element \(s\). 
We will denote this sub-coalgebra as \(⟨s⟩_{S}\), and call it \emph{principle sub-coalgebra} generated by \(s\). 
We sometimes omit the subscript \(S\) when it can be inferred from context or irrelevant.
Intuitively, we usually think of principle sub-coalgebra \(⟨s⟩_S\) as the sub-coalgebra that is formed by all the ``reachable state'' form the state \(s\).
This coalgebraic characterization of reachable state can allow us to avoid induction on the length of path from \(s\) to reach another state.

For all coalgebra \(S\) and a state \(s ∈ S\), principle sub-coalgebra \(⟨s⟩_S\) always exists and is unique, because sub-coalgebra of any coalgebra forms a complete lattice~\cite[theorem 6.4]{rutten_UniversalCoalgebraTheory_2000}; thus taking the meet of all the sub-coalgebra that contains \(s\) will yield \(⟨s⟩_S\).

Similar to sub-coalgebra, principle sub-coalgebra is also preserved under homomorphic image:
\begin{theorem}\label{thm:homo-img-preserve-principle-sub-coalg}
    Homomorphic image preserves principle sub-GKAT coalgebra. Specifically, given a homomorphism \(h: S → U\):
    \[h(⟨s⟩_S) = ⟨h(s)⟩_U\]
\end{theorem}

\begin{proof}
    We will need to show that \(h(⟨s⟩_{S})\) is the smallest sub-GKAT coalgebra of \(S\) that contain \(h(s)\). First by definition of image, \(h(s) ∈ h(⟨s⟩_{S})\); second by \cref{thm:hom-(pre)img-preserve-sub-coalg}, \(h(⟨S⟩_S) ⊑ U\).

    Finally, take any \(U' ⊑ U\) and \(h(s) ∈ U'\), recall that by \cref{thm:hom-(pre)img-preserve-sub-coalg}, \(h^{-1}(U') ⊑ S\). We can then derive that \(h(⟨s⟩_S) ⊑ U'\): 
    \begin{align*}
        h(s) ∈ U' 
        & ⟹ s ∈ h^{-1}(U') \\  
        & ⟹ ⟨s⟩_S ⊑ h^{-1}(U') & \text{definition of \(⟨s⟩_S\)}\\  
        & ⟹ h(⟨s⟩_S) ⊑ U' & \text{\cref{thm:hom-(pre)img-preserve-sub-coalg}}
    \end{align*}
    Hence \(h(⟨s⟩_S)\) is the smallest sub-GKAT coalgebra of \(U\) that contains \(h(s)\).
\end{proof}

@@

Given two \(F\)-coalgebra \(S\) and \(U\), the \emph{behavioral equivalence} between states in \(S\) and \(U\) can be computed by a notion called \emph{bisimulation}.
A relation \({∼} ⊆ S × U\) is called a \emph{bisimulation relation} if it forms an \(F\)-coalgebra: \[σ_{∼}: {∼} → F(∼),\] 
And its projections \(π₁: S × U → S\) and \(π₂: S × U → U\) are both homomorphisms:
\[
    \begin{tikzcd}
        S \ar[swap]{d}{σ_S}
            & ∼ \ar[swap]{l}{π₁} \ar{r}{π₂} \ar{d}{σ_{∼}} 
            & U \ar{d}{σ_T}\\  
        F(S) & F(∼) \ar{l}{F(π₁)} \ar[swap]{r}{F(π₂)} & F(T)
    \end{tikzcd}
\]
In a special case, when there exists a homomorphism \(h: S → U\), then we can simply pick \({∼} ⊆ S × U\) to be \(\{(s, h(s)) ∣ s ∈ S\}\), which gives us a bisimulation with the lift \(σ_{∼} ≜ σ_S × σ_U\).

\begin{corollary}
    Given two \(F\)-coalgebra \(S, U\) and a homomorphism \(h: S → U\), then all for all \(s\), \(⟦s⟧_{S} = ⟦h(s)⟧_{U}.\)
\end{corollary}

@@

    The condition \labelcref{itm:acc-condition} holds:
    \begin{align*}
        & δ_S(s, α) = \accept \\
        ⟺ & \norm(δ_S)(s, α) = \accept \\
        ⟺ & \norm(δ_∼)((s,t), α) = \accept \\ 
        ⟺ & \norm(δ_T)(t, α) = \accept \\
        ⟺ & δ_T(t, α) = \accept
    \end{align*}

    The condition~\labelcref{itm:transition-bisim} holds, by case analysis on the liveness of \(s'\) and \(t'\).
    First note that \(s'\) and \(t'\) has to be both live or both dead: because \(δ_S(s, α) = (s', p)\), then \(\norm(δ_S)(s', α)\) can either be rejection or \((s',p)\), and so is \(\norm(δ_T)(t', α)\). Finally because \(s ∼ t\), then
    \begin{align*}
        s' \text{ is live} 
        & ⟺ \norm(δ_S)(s, α) = (s', p) \\
        & ⟺ \norm(δ_T)(t, α) = (t', p) \\
        & ⟺ t' \text{ is live}.
    \end{align*}
    \begin{itemize}
        \item If both \(s'\) and \(t'\) are live, then \(s' ∼ t'\). By~\cref{thm:sub-coalg-preserve-bisim}, the bisimulation \(∼_{s', t'}\) is just \(∼\) restricted to \(⟨s'⟩\) and \(⟨t'⟩\).
        \item If both \(s'\) and \(t'\) are dead, then \(∼_{s', t'}\) can just be the singleton relation, according to~\cref{thm:bisim-between-dead}.
    \end{itemize}

@@

    The condition~\labelcref{itm:rej-or-dead} holds: 
    \begin{align*}
        & δ_S(s, α) \text{ rejects or transition to dead states} \\
        ⟺ & \norm(δ_S)(s, α) = \reject \\
        ⟺ & \norm(δ_T)(t, α) = \reject \\
        ⟺ & δ_T(t, α) \text{ rejects or transition to dead states}.
    \end{align*}
